Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
pass through section 7

@@

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
reduction can not take any real reduction steps and can only step into a term
itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
sufficient.  In fact, all the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
immediately discharged by Lemma~\ref{lemma:parnenf}.

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal form\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
  natural numbers as term variables. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}

@@

\scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
sentence elsewhere? I'd like to use the previous sentence as a conclusion for
this section.}
Finally, due to technical
details, our mechanized proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & = & = \\
    Renaming & 46 & =  & = \\
    Untyped reduction & 349 & = & = \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool, available from \url{https://github.com/AlDanial/cloc}.} for each file of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the extension to
normalization for open and closed terms from Section~\ref{sec:extension}.

Because the normalization and the consistency development share the same
syntactic and typing specification, they differ in only the categories
that are relevant to the logical relation. We use = as a marker that
the line count is the same.

Autosubst 2 takes our syntax specification in higher-order abstract
syntax and generates Coq syntax specification, renaming and
substitution functions, and lemmas and tactics that allow reasoning
about those functions. The auto-generated syntax file by Autosubst 2
and library header files are omitted from the figure.

Orthogonal to the development of our consistency and normalization
proof, we have also proven the syntactic soundness of our system through subject
reduction and progress. The syntactic soundness proof shares the
definition of renaming with our semantic soundness proof, which is
factored out as a separate file under the Renaming (common) category.
\scw{Should add this line count as another column, for comparison. I've done so,
but may need to update the numbers. }

\subsection{Artifacts Specific to Coq}
In this section, we discuss the Coq encoding of the definitions and proofs presented
in Section~\ref{sec:logrelproof} and the artifacts of this encoding that are
specific to Coq.

@@

In our proof presented in Section~\ref{sec:logreldep}, the relation
$[[R]]$ in \rref{I-Pi} requires Coq's impredicative \texttt{Prop} sort
since $[[R]]$ can later be instantiated into something defined in
terms of the logical relation itself (e.g. in
Lemma~\ref{lemma:piintroalt}). \scw{I'm having trouble connecting the rule in
the figure with rule \texttt{InterpExtFun} in the coq development. Would it make
sense to include the definition of ProdSpace? What is the Coq type of R?}
\yl{I think I might just inline the definition of ProdSpace and then
  rename the variables from the Coq code to match the ones from the paper}

In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
has type \texttt {tm -> Prop}.  However, if desired, we could consistently
replace the use of \texttt{Prop} with Coq's predicative sort
\texttt{Type}. This variation of the definition could be used to define the
interpretation for any \emph{finite} number of universes. The use of
\texttt{Type} becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation defined in
Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
levels. Therefore, the one feature of \lang{} that truly requires impredicativity
is its countable universe hierarchy. \scw{I'm not sure the last part of this sentence
adds much: ``Although, lemmas such as the
admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only simply
our mechanization, but also give better intuition on what our logical relation
means.'' Yes, we need impredicativity to define these lemmas, but without impredicativity
our logical relation means something else.}

@@

Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
are equal, it suffices to show the extensional property that $\forall
x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
Prop}, a predicate over \lang{} terms.
In axiom-free Coq, predicates do not come with the extensionality
property. Given two predicates $P$ and $Q$, we cannot conclude that
$P = Q$ when given a proof of $\forall
x, P(x) \iff Q(x)$. While
predicate extensionality is not necessary in our development, we
assume it as an axiom since it is known to be consistent with Coq's
metatheory and helps bridge the gap between our mechanization and the
proofs we present.

@@

\subsection{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, though instead of
defining custom tactics, we rely mostly on off-the-shelf tools such as
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.

We use the Autosubst 2 framework to specify our syntax in HOAS and
produce Coq syntax files in de Bruijn representation. Additionally,
Autosubst 2 provides a powerful tactic \texttt{asimpl} that
can be used to prove the equivalence of two terms constructed using
the primitive operators provided by the framework. This greatly
simplifies the reasoning about substitution in our development as
almost all substitution related properties about the syntax are
immediately discharged by \texttt{asimpl}.

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also gives useful documentation for readers
who wish to understand the mechanized proof. Although automation performs
extensive search, it does not use lemmas or invert derivations that are not
specified in the \texttt{use} or \texttt{inv} flag.

@@

The soundness of the algorithm is immediate. The completeness of the
algorithm is justified the confluence property of the untyped
reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
f1]]$ but $[[f0]]$ is syntactically distinct from $[[f1]]$. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
Because $[[f0]]$ and $[[f1]]$ are both in
$\beta\eta$-normal forms and can only reduce to themselves, we must
have $[[f0 = f1]]$, which is a contradiction.

@@

However, % with a little more effort,
it is possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}.

>>>>

\scw{Need to acknowledge the lack of type annotation on abstractions.}

@@

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
reduction cannot take any real reduction steps and can only step into a term
itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
sufficient.  In fact, all the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
immediately discharged by Lemma~\ref{lemma:parnenf}.

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
  natural numbers as term variables. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}

@@

\scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
sentence elsewhere? I'd like to use the previous sentence as a conclusion for
this section.}
Finally, due to technical
details, our mechanized proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ |l |  c  | c | c| }
    \hline
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    Syntactic typing         &  83 & =   & = \\
    Untyped reduction        & 344 & =   & = \\
    Neutral and normal forms & -   & 273 & - \\
    Logical relation         & 338 & 430 & - \\
    Semantic soundness       & 190 & 214 & - \\
    Syntactic soundness      & -   & -   & 629 \\
    \hline
    Total                    & 955 & 1344 & 1056 \\
    \hline
  \end{tabular}
  \end{minipage}
  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
left. The marker - indicates the file does not
contribute to the total.}
  \label{fig:linecount}
\end{figure}

To demonstrate the scale of our proof scripts, 
Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool, available from
  \url{https://github.com/AlDanial/cloc}.} for each file of our development,
including the base consistency proof from Section~\ref{sec:logreldep} and
\ref{sec:logrelproof} and the extension to $\beta$-normalization for open and
closed terms from Section~\ref{sec:extension}.  For comparison, we have also
proven syntactic type soundness through subject
reduction\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.

The Autosubst 2 tool takes our 13 line syntax specification, written in
higher-order abstract syntax, and generates the Coq syntax specification,
renaming and substitution functions, and lemmas and tactics that allow
reasoning about those functions. The auto-generated syntax file (291 LOC)
and other Autosubst library files are are not included in the figure.

\subsection{Axioms}
Our Coq development assumes two axioms: functional extensionality 
and propositional extensionality. The former is also required by 
the Autosubst 2 libraries. Both axioms are known to be consistent 
with Coq's metatheory.

Furthermore, these axioms also bridge the gap between our mechanization 
and our informal proofs.

For example, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$ are
equal, it suffices to show the extensional property that
$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.

In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm
  -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
not come with the extensionality property. Given two predicates $P$ and $Q$,
we cannot conclude that $P = Q$ when given a proof of
$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of the 
axiom of functional extensionality.

\subsection{Encoding the logical relation in Coq}
We next discuss specific details of the Coq encoding of the logical relation
presented in Section~\ref{sec:logreldep}.

First, this definition requires the use of Coq's impredicatve \texttt{Prop}
sort. In particular, the relation $[[R]]$ in \rref{I-Pi} requires the use of
impredicativity because $[[R]]$ must later be instantiated into something
defined in terms of the logical relation itself (e.g. in
Lemma~\ref{lemma:piintroalt}).

@@

In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.

However, if desired, we could consistently replace the use of \texttt{Prop}
with Coq's predicative sort \texttt{Type} in the definition of
$[[Interp I i A S]]$. This alternative definition could be part of
the interpretation for any \emph{finite} number of universes. The use of
\texttt{Type} becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation defined in
Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
levels. Therefore, the one feature of \lang{} that truly requires
impredicativity is its countable universe hierarchy. \scw{I'm not sure the
  last part of this sentence adds much: ``Although, lemmas such as the
  admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only
  simply our mechanization, but also give better intuition on what our logical
  relation means.'' Yes, we need impredicativity to define these lemmas, but
  without impredicativity our logical relation means something else.}

@@

@@

\subsection{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, supported by the tools
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.

We use the Autosubst 2 framework to produce Coq syntax files based on a de
Bruijn representation of variable binding and capture-avoiding substitution.
In addition to these generated definitions, Autosubst 2 provides a powerful
tactic \texttt{asimpl} that can be used to prove the equivalence of two terms
constructed using the primitive operators provided by the framework. This
tactic simplifies the reasoning about substitution as many
substitution-related properties about syntax are immediately discharged by
\texttt{asimpl}.

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also provides useful documentation for
readers who wish to understand the mechanized proof. Although automation
performs extensive search, it does not use lemmas or invert derivations that
are not specified in the \texttt{use} or \texttt{inv} flags.

@@

The soundness of the algorithm is immediate. The completeness of the
algorithm is justified by the confluence property of the untyped
reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
f1]]$ but $[[f0]]$ is syntactically distinct from $[[f1]]$. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
Because $[[f0]]$ and $[[f1]]$ are both in
$\beta\eta$-normal forms and can only reduce to themselves, we must
have $[[f0 = f1]]$, which is a contradiction.

@@

However, % with a little more effort,
it should be possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}.
